The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2012-08-22 script/en/chapters/primes.tex
Some small corrections in the CT Script
CT Script: Many minor corrections and updates (e.g. 4 levels of titles)

\newpage
% \hypertarget{Kapitel_2}{}
\chapter{Prime Numbers}
\label{Label_Kapitel_Primes}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007, January 2010)

@@

Prime numbers are particularly important for 3 reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (digital
\index{Signature!digital} signature, elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{Performance} of computers) and leads
to new calculation methods on many computers \\
(see also: \url{http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millennia.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also
explain what is known about the distribution (density, number of prime numbers
in particular intervals) of prime numbers and how prime number tests work.

@@

\hypertarget{euclid}{}
\paragraph{Euclid's proof by contradiction}
\index{Euclid's proof by contradiction}\index{Proof by contradiction}
goes as follows:

@@

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsection{The 20+ largest known primes (as of July 2009)}  % Eyecatcher_neue-Mersenne
\label{RecordPrimes}
\index{Prime number!records}

@@

	4  & $2^{32,582,657}-1$ &  9,808,358 & 2006 & Mersenne, 44th known \\
	5  & $2^{30,402,457}-1$ &  9,152,052 & 2005 & Mersenne, 43rd known \\
	6  & $2^{25,964,951}-1$ &  7,816,230 & 2005 & Mersenne, 42nd known \\
	7  & $2^{24,036,583}-1$ &  7,235,733 & 2004 & Mersenne, 41st known \\
	8  & $2^{20,996,011}-1$ &  6,320,430 & 2003 & Mersenne, 40th known \\
	9  & $2^{13,466,917}-1$ &  4,053,946 & 2001 & Mersenne, M-39 \\
	10 & $19,249 \cdot 2^{13,018,586}+1$ & 3,918,990 & 2007 & Generalized Mersenne\footnotemark \\
	11 & $27,653 \cdot 2^{9,167,433}+1$ & 2,759,677 & 2005 & Generalized Mersenne \\

@@

\hline
\end{tabular}
\caption{The 20+ largest known primes and its particular number types
         (as of July 2009)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\footnotetext{\index{Number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB) (\url{http://www.seventeenorbust.com})
% {\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its
task sometime.

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before 
\index{Lucas, Edouard} Edouard Lucas (1842-1891) proved that this number 
is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson Cole\index{Cole, Frank Nelson}\footnote{%
Frank Nelson Cole, American mathematician, Sep. 20, 1861 $-$ May 26, 1926.}
showed in 1903 which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorization 
\index{Factorization} (expression as a product of prime factors)\footnote{%
  Using CrypTool\index{CrypTool} you can factorize numbers in the 
  following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
  \textbackslash{} Factorization of a Number}. \\
  CrypTool can factorize in a reasonable time numbers no longer than 250 bit.
  Numbers bigger than 1024 bits are currently not accepted by CrypTool. \\
  The current factorization records are listed in chapter \ref{nt:NoteFactorization}.
  \index{Factorization!factoring records}
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
The following page from Landon Curt Noll\index{Noll, Landon Curt} contains 
all Mersenne primes including its date of discovery and its value as number 
and as word:
      \url{http://www.isthe.com/chongo/tech/math/prime/mersenne.html} \\
Also see:
      \url{http://www.utm.edu/}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217;\\
4,253; ~ 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; \\
132,049; ~ 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377;\\
6,972,593; ~ 13,466,917; ~ 20,996,011; ~ 24,036,583; ~ 25,964,951; ~ 30,402,457;\\
32,582,657; ~ 37,156,667; ~ 43,112,609; ~ 42,643,801.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus    % Eyecatcher_neue-Mersenne
$47$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}.

@@

\vskip +25 pt
For the first 39            % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 40th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely\footnote{%
The current status of the check can be found at:
      \url{http://www.mersenne.org/status.htm}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{Prime number!test}. }.

\vskip +25 pt
\paragraph{M-37 -- January 1998}\index{Mersenne!prime number!M-37}\mbox{}

@@

\vskip +25 pt
\paragraph{M-38 -- June 1999}\index{Mersenne!prime number!M-38}\mbox{}

@@

\vskip +25 pt
\hypertarget{M-39}{}
\paragraph{M-39 -- December 2001}%
\index{Mersenne!prime number!M-39}\mbox{}

@@

Right now (July 2009) all prime exponents smaller than $ 18,000,989 $ have been
tested and double-checked\footnote{%
See home page of the GIMPS project:
      \url{http://www.mersenne.org/report_milestones}.
                                  }:
So we can
be certain, that this is really the 39th Mersenne prime number and that
there are no smaller undiscovered Mersenne primes (it is common usage to use
the notation M-nn not until it is proven, that the nn-th known Mersenne prime
is really the nn-th Mersenne prime).

%\vskip +15 pt
%\paragraph{Mxxxxxxxxxx -- June 2003 -- M-40 ?}%
%\index{Mersenne!prime number!M-40}\mbox{}
%\vskip +10 pt

@@

% --------------------------------------------------------------------------
\section[Prime number tests]{Prime number tests\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply the
    tests of Fermat and of Miller-Rabin:
    See learning units 3.2 and 3.3, pages 3-11/11.\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\label{primality_tests}   % chap. 3.5
\index{Prime number!test}

@@

If we look for the prime factors in order to decide whether a number is prime,
then the search takes too long, if even the smallest prime factor is enormous.
Factorizing numbers using systematic computational
division or using the \hyperlink{SieveEratosthenes01}{sieve of Eratosthenes} 
\index{Eratosthenes!sieve} is only feasible using current computers for
numbers with up to around $20$ digits in the decimal system.
The biggest number factorized into its 2 almost equal prime factors 
has 200 digits 
(see \hyperlink{RSA-200-chap3}{RSA-200} in chapter \ref{nt:NoteFactorization}).
% be_2005_UPDATEN_if-new-factorization-record-appears

@@

\vskip +25 pt
\paragraph{Pseudo prime numbers}%
\index{Prime number!pseudo prime} \index{Number!pseudo prime}%
\hypertarget{HT-Pseudoprimenumber01}{}\label{L-Pseudoprimenumber01}%
\mbox{}
\vskip +10 pt
Numbers n that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$
but are not prime are called {\em pseudo prime numbers}
(i.e. the exponent is not a prime). 
The first pseudo prime number is $$ 341 = 11 \cdot 31 .$$

\vskip +25 pt
\paragraph{Carmichael numbers}%
\index{Number!Carmichael}%
\hypertarget{HT-Carmichael-number01}{}\label{L-Carmichael-number01}%
\mbox{}
\vskip +10 pt
There are pseudo prime numbers n that pass the Fermat test 
$$ a^{n-1} \equiv 1 \;{\rm mod}\; n $$
with all bases a which are relatively prime to n [$ gcd (a,n) = 1 $],
despite these numbers n are not prime: These numbers are called
{\em Carmichael numbers}.
The first of these is $$ 561 = 3 \cdot 11 \cdot 17 .$$

@@

\vskip +25 pt
\paragraph{Strong pseudo prime numbers}%
\index{Prime number!strong pseudo prime}\index{Number!strong pseudo prime}%
\hypertarget{HT-Strongpseudoprimenumber01}{}\label{L-Strongpseudoprimenumber01}%
\mbox{}
\vskip +10 pt
A stronger test is provided by\index{Miller, Gary L.}\index{Rabin, Michael O.}
Miller/Rabin\footnote{
In 1976 an efficient probabilistic primality test was published by Prof. Rabin, based on a number theoretic result of Prof. Miller from the year before. \\
Prof. Miller worked at the Carnegie-Mellon University, School of Computer
Science. Prof. Rabin, born in 1931, worked at the Harvard and Hebrew University.
}%
: it is only passed by so-called {\em strong pseudo prime numbers}. 
Again, there are strong pseudo prime numbers that are not primes, but this is
much less often the case than for (simple) pseudo prime numbers or for 
Carmichael numbers. The smallest strong pseudo prime number base $2$ is 
$$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

% --------------------------------------------------------------------------
\vskip +30 pt
\section{Overview special number types and the search for a formula for primes}\label{spezialzahlentypen}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Catalan's conjecture]{Catalan's conjecture\footnotemark}
    \footnotetext{%
Eugene Charles Catalan, Belgian mathematician, May 5, 1814$-$Feb 14, 1894.\\
After him the so-called {\em Catalan numbers}
$A(n) = (1 / (n+1) ) * (2n)! / (n!)^2 $ \\
$=  1, 2, 5, 14, 42, 132, 429, 1430, 
4862, 16796, 58786, 208012, 742900, 2674440, 9694845, ... $
are named.
    }
\nopagebreak
Catalan \index{Catalan Eugene}  \index{Number!Catalan}
conjectured that $ C_4 \;$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
%% \begin{sloppypar}
\noindent (see \url{http://www.utm.edu/research/primes/mersenne.shtml}
           under ``Conjectures and Un\-sol\-ved Problems'').
%% \end{sloppypar}

    This sequence is also defined recursively and increases extremely quickly.
Does it only consist of primes?
$$
\begin{array}{lll}
C(0) = 2 & & \mapsto {\rm \:prime}\\
C(1) = 2^2 - 1 &    = 3 & \mapsto {\rm \:prime}\\
C(2) = 2^3 - 1 &    = 7 & \mapsto {\rm \:prime} \\
C(3) = 2^7 - 1 &    = 127& \mapsto {\rm \:prime} \\
C(4) = 2^{127} - 1 &      = 170, 141, 183, 460, 469, 231, 731, 687, 303, 715,
884, 105, 727 & \mapsto {\rm \:prime} \\
\end{array} 
$$
It is not (yet) known whether $C_5$ and all higher elements are prime, but
this is not very likely.
In any case, it has not been proved that this formula delivers only primes.

@@

\vskip +15 pt
\paragraph{The value of the $n$-th prime $P(n)$}%
\index{P(n)}%
\mbox{}
\vskip +10 pt

@@

\vskip +15 pt
\paragraph{The number of prime numbers $PI(x)$}%
\index{PI(x)}%
\mbox{}
\vskip +10 pt
The definition is similar for the number of prime numbers $PI(x)$
that do not exceed the maximum value $x$:

@@

%----------------------------------------
\vskip +15 pt
\paragraph{Sieve of Eratosthenes}\mbox{}
\hypertarget{SieveEratosthenes01}{}
\index{Eratosthenes!sieve}

@@

\paragraph{Further interesting topics regarding prime numbers}
\mbox{}\\
This chapter \ref{Label_Kapitel_Primes} didn't consider other number theory topics
such as divisibility rules, modulus calculation, modular inverses, modular
powers, modular roots, Chinese remainder theorem, Euler Phi function or
perfect numbers.
Some of these topics are considered in the \hyperlink{Chapter_ElementaryNT} 
{{\bf next chapter}} (chapter \ref{Chapter_ElementaryNT}).

@@

     \hline
     \end{tabular}
     \caption{Arithmetic prime number sequences with minimal difference (as of Jan. 2010)}
     \label{Smallest AP-k with minimal difference}
     \end{center}
     \end{table}
     \clearpage

     If someone wants to discover such sequences he should consider the
     following thought.
     The length of a sequence determines the minimal common distance
     between the single primes of the sequence. Given a sequence with 6 
     elements the distance between them has to be 30 or a multiple of 30.
     The number 30 results as the product of all primes smaller than the
     length of the sequence. So its the product of all primes smaller than 6:
     $ 6\# = 5\# = 2 * 3 * 5 = 30 $.
     It is $ 10\# = 7\# = 2 ∑ 3 ∑ 5 ∑ 7 = 210$.
     If you look for a sequence with 15 elements, then the common
     distance is at least $15\# = 13\# = 2 * 3 * 5 * 7 * 11 * 13 = 30,030$.

@@

\vskip +25 pt
\hypertarget{HT-GoogleRecruitment2004}{}
\paragraph{Recruitment at Google in 2004}%
\index{Google!Recruitment}%
\label{HT-GoogleRecruitment2004}%
\mbox{}
\vskip +10 pt

@@

\vskip +25 pt
\hypertarget{HT-Movie-Contact01}{}
\paragraph{Contact [Directed by Robert Zemeckis, 1997] -- Primes helping to 
           contact aliens}
\index{Movies}\index{Zemeckis 1997}
\label{HT-Movie-Contact01}%
\mbox{}
\vskip +10 pt

>>>>

\newpage
% \hypertarget{Kapitel_2}{}
\chapter{Prime Numbers}
\label{Label_Kapitel_Primes}
(Bernhard Esslinger, May 1999; Updates Nov. 2000, Dec. 2001, June 2003, May 2005, March 2006, June 2007, January 2010, August 2012)

@@

Prime numbers are particularly important for three reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (digital
\index{Signature!digital} signature, elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{Performance} of computers) and leads
to new calculation methods on many computers \\
(see also: \url{http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millennia.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also
explain what is known about the distribution (density, number of prime numbers
in particular intervals) of prime numbers and how prime number tests work.

@@

\hypertarget{euclid}{}
\paragraph*{Euclid's proof by contradiction}
\index{Euclid's proof by contradiction}\index{Proof by contradiction}
goes as follows:

@@

% --------------------------------------------------------------------------
\hypertarget{RecordPrimes}{}
\subsection{The 20+ largest known primes (as of July 2012)}  % Eyecatcher_neue-Mersenne
\label{RecordPrimes}
\index{Prime number!records}

@@

	4  & $2^{32,582,657}-1$ &  9,808,358 & 2006 & Mersenne, 44th known \\
	5  & $2^{30,402,457}-1$ &  9,152,052 & 2005 & Mersenne, 43rd known \\
	6  & $2^{25,964,951}-1$ &  7,816,230 & 2005 & Mersenne, 42nd known \\
	7  & $2^{24,036,583}-1$ &  7,235,733 & 2004 & Mersenne, M-41 \\
	8  & $2^{20,996,011}-1$ &  6,320,430 & 2003 & Mersenne, M-40 \\
	9  & $2^{13,466,917}-1$ &  4,053,946 & 2001 & Mersenne, M-39 \\
	10 & $19,249 \cdot 2^{13,018,586}+1$ & 3,918,990 & 2007 & Generalized Mersenne\footnotemark \\
	11 & $27,653 \cdot 2^{9,167,433}+1$ & 2,759,677 & 2005 & Generalized Mersenne \\

@@

\hline
\end{tabular}
\caption{The 20+ largest known primes and its particular number types
         (as of July 2012)}    % Eyecatcher_neue-Mersenne
\label{L_n_Largest_Known-Primes}
\end{center}
\end{table} 
\footnotetext{\index{Number!Sierpinski}\index{Seventeen or Bust SoB}%
This number was found within the distributed computing project
``Seventeen or Bust'' (SoB) (\url{http://www.seventeenorbust.com})
% {\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}
at March 26, 2007. While the well known \hyperlink{GIMPS-project}{GIMPS project} (chapter~\ref{zahlentyp_mersenne}) searches for bigger and bigger of the infinitely many primes, there is a chance, that the SoB project could have been completed its
task sometime.

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before 
\index{Lucas, Edouard} Edouard Lucas (1842-1891) proved that this number 
is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson Cole\index{Cole, Frank Nelson}\footnote{%
Frank Nelson Cole, American mathematician, Sep. 20, 1861 $-$ May 26, 1926.}
showed in 1903 which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorization 
\index{Factorization} (expression as a product of prime factors)\footnote{%
  Using CrypTool 1\index{CrypTool 1} you can factorize numbers in the 
  following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
  \textbackslash{} Factorization of a Number}. \\
  CrypTool 1 can factorize in a reasonable time numbers no longer than 250 bit.
  Numbers bigger than 1024 bits are currently not accepted by CrypTool 1. \\
  The current factorization records are listed in chapter \ref{nt:NoteFactorization}.
  \index{Factorization!factoring records}
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!prime number} have the exponents\footnote{%
The following page from Landon Curt Noll\index{Noll, Landon Curt} contains 
all Mersenne primes including its date of discovery and its value as number 
and as word:
      \url{http://www.isthe.com/chongo/tech/math/prime/mersenne.html} \\
Also see:
      \url{http://www.utm.edu/}.
                             } 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217;\\
4,253; ~ 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; \\
132,049; ~ 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377;\\
6,972,593; ~ 13,466,917; ~ 20,996,011; ~ 24,036,583; ~ 25,964,951; ~ 30,402,457;\\
32,582,657; ~ 37,156,667; ~ 42,643,801, ~ 43,112,609.
% be_2005_UPDATEN_if-new-mersenne-prime-appears          ~ xxx,xxx,xxx.
\end{array}
$$
Thus    % Eyecatcher_neue-Mersenne
$47$            % be_2005_UPDATEN_if-new-mersenne-prime-appears
Mersenne prime numbers are currently known%
\index{Prime number!Mersenne}\index{Mersenne!prime number}.

@@

\vskip +10 pt
For the first 41            % be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
Mersenne prime numbers we know that this list is complete.
The exponents until the 47th     % be_2005_UPDATEN_if-new-mersenne-prime-appear    % Eyecatcher_neue-Mersennes
Mersenne prime number have not yet been checked completely\footnote{%
The current status of the check can be found at:
      \url{http://www.mersenne.org/status.htm}.\\
Hints, how the primality of a number can be checked, are in chapter
\ref{primality_tests}, prime number tests\index{Prime number!test}. }.

% be_2005_UPDATEN_if-new-mersenne-prime-appears    % Eyecatcher_neue-Mersenne
Right now (July 2012) all prime exponents smaller than $ 24,212,801 $ have been
tested and double-checked\footnote{%
See home page of the GIMPS project:
      \url{http://www.mersenne.org/report_milestones}.
                                  }:
So we can
be certain, that this is really the 41th Mersenne prime number and that
there are no smaller undiscovered Mersenne primes (it is common usage to use
the notation M-nn not until it is proven, that the nn-th known Mersenne prime
is really the nn-th Mersenne prime). Here are some examples in more detail:

\vskip +15 pt
\paragraph*{M-37 -- January 1998}\index{Mersenne!prime number!M-37}\mbox{}

@@

\vskip +25 pt
\paragraph*{M-38 -- June 1999}\index{Mersenne!prime number!M-38}\mbox{}

@@

\vskip +25 pt
\hypertarget{M-39}{}
\paragraph*{M-39 -- December 2001}%
\index{Mersenne!prime number!M-39}\mbox{}

@@

%\vskip +15 pt
%\paragraph*{Mxxxxxxxxxx -- June 2003 -- M-40 ?}%
%\index{Mersenne!prime number!M-40}\mbox{}
%\vskip +10 pt

@@

% --------------------------------------------------------------------------
\section[Prime number tests]{Prime number tests\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply the
    tests of Fermat and of Miller-Rabin:
    See learning units 3.2 and 3.3, pages 3-11/11.\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See also appendix \ref{s:appendix-Learn-NT}.
}
\label{primality_tests}   % chap. 3.5
\index{Prime number!test}

@@

If we look for the prime factors in order to decide whether a number is prime,
then the search takes too long, if even the smallest prime factor is enormous.
Factorizing numbers using systematic computational
division or using the \hyperlink{SieveEratosthenes01}{sieve of Eratosthenes} 
\index{Eratosthenes!sieve} is only feasible using current computers for
numbers with up to around $20$ digits in the decimal system.
The biggest number factorized into its 2 almost equal prime factors 
has 232 digits 
(see \hyperlink{RSA-768-chap3}{RSA-768} in chapter \ref{nt:NoteFactorization}).
% be_2005_UPDATEN_if-new-factorization-record-appears

@@

\vskip +25 pt
\paragraph*{Pseudo prime numbers}%
\index{Prime number!pseudo prime} \index{Number!pseudo prime}%
\hypertarget{HT-Pseudoprimenumber01}{}\label{L-Pseudoprimenumber01}%
\mbox{}
\vskip +10 pt
Numbers n that have the property $$ 2^n \equiv 2 \;{\rm mod}\; n $$
but are not prime are called {\em pseudo prime numbers}
(i.e. the exponent is not a prime). 
The first pseudo prime number is $$ 341 = 11 \cdot 31 .$$

\vskip +25 pt
\paragraph*{Carmichael numbers}%
\index{Number!Carmichael}%
\hypertarget{HT-Carmichael-number01}{}\label{L-Carmichael-number01}%
\mbox{}
\vskip +10 pt
There are pseudo prime numbers n that pass the Fermat test 
$$ a^{n-1} \equiv 1 \;{\rm mod}\; n $$
with all bases a which are relatively prime to n [$ gcd (a,n) = 1 $],
despite these numbers n are not prime: These numbers are called
{\em Carmichael numbers}.
The first of these is $$ 561 = 3 \cdot 11 \cdot 17 .$$

@@

\vskip +25 pt
\paragraph*{Strong pseudo prime numbers}%
\index{Prime number!strong pseudo prime}\index{Number!strong pseudo prime}%
\hypertarget{HT-Strongpseudoprimenumber01}{}\label{L-Strongpseudoprimenumber01}%
\mbox{}
\vskip +10 pt
A stronger test is provided by\index{Miller, Gary L.}\index{Rabin, Michael O.}
Miller/Rabin\footnote{
In 1976 an efficient probabilistic primality test was published by Prof. Rabin, based on a number theoretic result of Prof. Miller from the year before. \\
Prof. Miller worked at the Carnegie-Mellon University, School of Computer
Science. Prof. Rabin, born in 1931, worked at the Harvard and Hebrew University.
}%
: it is only passed by so-called {\em strong pseudo prime numbers}. 
Again, there are strong pseudo prime numbers that are not primes, but this is
much less often the case than for (simple) pseudo prime numbers or for 
Carmichael numbers. The smallest strong pseudo prime number base $2$ is 
$$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

% --------------------------------------------------------------------------
\vskip +30 pt
\section{Special number types and the search for a formula for primes}\label{spezialzahlentypen}
\index{Prime number!formula}
There are currently no useful, open (i.e. not recursive) formulae known that
only deliver prime numbers (recursive means that in order to calculate the
function the same function is used with a smaller variable). Mathematicians
would be happy if they could find a formula that leaves gaps (i.e. does not
deliver all prime numbers) but does not deliver any composite (non-prime)
numbers.

@@

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Catalan's Mersenne conjecture]{Catalan's Mersenne conjecture\footnotemark}
    \footnotetext{%
Eugene Charles Catalan, Belgian mathematician, May 5, 1814$-$Feb 14, 1894.\\
After him, the so-called {\em Catalan numbers}
$A(n) = (1 / (n+1) ) * (2n)! / (n!)^2 $ \\
$=  1, 2, 5, 14, 42, 132, 429, 1430, 
4862, 16796, 58786, 208012, 742900, 2674440, 9694845, ... $
are named.
    }
\nopagebreak
Catalan \index{Catalan Eugene}  \index{Number!Catalan}
conjectured\footnote{%
See \url{http://www.utm.edu/research/primes/mersenne.shtml}
           under ``Conjectures and Un\-sol\-ved Problems''.
         }
 that $ C_4 \;$ and any further term in this sequence is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$

\noindent This sequence is defined recursively and increases extremely quickly.
Does it only consist of primes?

$$
\begin{array}{lll}
C(0) = 2 & & \mapsto {\rm \:prime}\\
C(1) = 2^2 - 1 &    = 3 & \mapsto {\rm \:prime}\\
C(2) = 2^3 - 1 &    = 7 & \mapsto {\rm \:prime} \\
C(3) = 2^7 - 1 &    = 127& \mapsto {\rm \:prime} \\
C(4) = 2^{127} - 1 &      = 170, 141, 183, 460, 469, 231, 731, 687, 303, 715,
884, 105, 727 & \mapsto {\rm \:prime} \\
\end{array} 
$$

It is not (yet) known whether $C_5 = 2^{C_4} - 1$ and all higher elements are prime.
In any case, it has not been proved that this formula delivers only primes.

\noindent It seems very unlikely that $C_5$ (or many of the larger terms) would be prime.

\noindent So this could be another example of Guy's ``law of small numbers''
\index{Guy's law of small numbers}\footnote{%
\url{http://primes.utm.edu/glossary/page.php?sort=LawOfSmall}
               }.

% --------------------------------------------------------------------------
\vskip +10 pt
\subsection[Double Mersenne primes]
           {Double Mersenne primes}

The above Catalan-Mersenne numbers are a subset of the double Mersenne
primes\footnote{%
\url{http://en.wikipedia.org/wiki/Catalan%E2%80%93Mersenne_number}
          }.
A double Mersenne prime is a Mersenne prime of the form
$$ M_{M_p} = 2 ^ {2^p - 1} -1 $$
% $ M_{M_p} = 2 ^ {2^p - 1} -1 $ 
where p is a Mersenne prime exponent and $M_p$ ist a prime Mersenne number.

The first values of p for which $M_p$ is prime are p = 2, 3, 5, 7, 13, 17, 19, 31, 61, 89 (see \hyperlink{MersenneNumbers01}{above}).

$M_{M_p}$ is known to be prime for p = 2, 3, 5, 7.
For p = 13, 17, 19, and 31, the corresponding double Mersenne numbers are not prime.

The next candidate for the next double Mersenne prime is $ M_{M_{61}} = 2^{2305843009213693951} - 1$.
Being approximately $ 1.695 * 10^{694,127,911,065,419,641} $, this number --- like
$C_5$ --- is far too large for any currently known primality test to be
successfully applied.

@@

\vskip +15 pt
\paragraph*{The value of the $n$-th prime $P(n)$}%
\index{P(n)}%
\mbox{}
\vskip +10 pt

@@

\vskip +15 pt
\paragraph*{The number of prime numbers $PI(x)$}%
\index{PI(x)}%
\mbox{}
\vskip +10 pt
The definition is similar for the number of prime numbers $PI(x)$
that do not exceed the maximum value $x$:

@@

%----------------------------------------
\vskip +15 pt
\paragraph*{Sieve of Eratosthenes}\mbox{}
\hypertarget{SieveEratosthenes01}{}
\index{Eratosthenes!sieve}

@@

\paragraph*{Further interesting topics regarding prime numbers}
\mbox{}\\
This chapter \ref{Label_Kapitel_Primes} didn't consider other number theory topics
such as divisibility rules, modulus calculation, modular inverses, modular
powers, modular roots, Chinese remainder theorem, Euler Phi function or
perfect numbers.
Some of these topics are considered in the \hyperlink{Chapter_ElementaryNT} 
{{\bf next chapter}} (chapter \ref{Chapter_ElementaryNT}).

@@

     \hline
     \end{tabular}
     \caption{Arithmetic prime number sequences with minimal difference (as of Aug. 2012)}
     \label{Smallest AP-k with minimal difference}
     \end{center}
     \end{table}
     \clearpage

     If someone wants to discover such sequences he should consider the
     following thought.
     The length of a sequence determines the minimal common distance
     between the single primes of the sequence. Given a sequence with 6 
     elements the distance between them has to be 30 or a multiple of 30.
     The number 30 results as the product of all primes smaller than the
     length of the sequence. So its the product of all primes smaller than 6:
     $ 6\# = 5\# = 2 * 3 * 5 = 30 $.
     It is $ 10\# = 7\# = 2 ∑ 3 ∑ 5 ∑ 7 = 210$.
     If you look for a sequence with 15 elements, then the common
     distance is at least $15\# = 13\# = 2 * 3 * 5 * 7 * 11 * 13 = 30,030$.
     \index{Primzahl!k Primorial} \index{Primzahl!k\#}

@@

\vskip +25 pt
\hypertarget{HT-GoogleRecruitment2004}{}
\paragraph*{Recruitment at Google in 2004}%
\index{Google!Recruitment}%
\label{HT-GoogleRecruitment2004}%
\mbox{}
\vskip +10 pt

@@

\vskip +25 pt
\hypertarget{HT-Movie-Contact01}{}
\paragraph*{Contact [Directed by Robert Zemeckis, 1997] -- Primes helping to 
           contact aliens}
\index{Movies}\index{Zemeckis 1997}
\label{HT-Movie-Contact01}%
\mbox{}
\vskip +10 pt
